---
title: "Tidying Data Lab"
author: "Anthony Tetreault"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

1.  The built in billboard dataset is not tidy. Describe why it is not tidy and then tidy the dataset.

```{r}
# This dataset is not tidy because the columns are values of a variable, weeks, not each variable has its own column.
billboard %>% 
    pivot_longer(
        cols = starts_with("wk"),
        names_to = "week",
        names_prefix = "wk",
        names_transform = list(week = as.integer),
        values_drop_na = TRUE
    ) %>% 
        mutate(date = date.entered + weeks(week)) %>% 
            arrange(artist, track, week)
```

2.  Tidy the "fish_encounters" dataset of fish spotting by monitoring stations. Make the NA into 0 using the option "values_fill = list(seen = 0)"

```{r}
fish_encounters %>% 
    pivot_wider(
        names_from = "station",
        values_from = "seen",
        values_fill = list(seen = 0)
    )
```

3.  Import the flowers1 dataset. Tidy and pivot the data. Hint: use "read_csv2()" to read in the dataset 

```{r}
flowers1 <- read_csv2("./flowers1.csv")

flowers1 %>% 
    pivot_wider(
        names_from = "Variable",
        values_from = "Value"
    )
```

4.  Import the flowers2 dataset. Tidy the dataset by turning the one column into 3 separate columns

```{r}
flowers2 <- read_csv2("./flowers2.csv")

flowers2 %>% 
    separate_wider_delim(
        cols = "Flowers/Intensity",
        delim = "/",
        names = c("Flowers", "Intensity")
    ) %>% select(Time, Flowers, Intensity)
```

5.  In the following dataset, turn the implicit missing values to explicit 

```{r}
output <- tibble(
      treatment   = c("a", "b", "a", "c", "b"),
      gender   = factor(c("M", "F", "F", "M", "M"), levels = c("M", "F", "O")),
      return = c(1.5, 0.75,  0.5,  1.8,  NA)
    )
output %>% 
    complete(treatment, gender)
```

6.  Import the weather dataset as weather. Use "pivot_longer()" to put the days all in one column, then use "pivot_wider" to separate tmax and tmin into separate columns. Print the summary of the final resulting dataset 

```{r}
weather <- read_csv("./weather.csv")

weather %>% 
    pivot_longer(
        cols = starts_with("d"),
        names_to = "day",
        names_prefix = "d",
        values_drop_na = TRUE
    ) %>% 
        pivot_wider(
            names_from = "element",
            values_from = "value"
        ) %>% 
            summary()
```

7.  Load the built in "anscombe" data frame and use "pivot_longer()" to separate all the x and y columns and categorize them into 4 sets.

```{r}
anscombe %>% 
     pivot_longer( 
         cols = starts_with(c("x","y")),
         names_to = c(".value", "set"),
         names_pattern = "(.)(.)"
     )
```

