---
title: "Practice Tidying Data"
author: "Anthony Tetreault"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# Tidying Data

1.  In the following data set, turn the implicit missing values to explicit.

```{r}
output <- tibble(
      treatment   = c("a", "b", "a", "c", "b"),
      gender   = factor(c("M", "F", "F", "M", "M"), levels = c("M", "F", "O")),
      return = c(1.5, 0.75,  0.5,  1.8,  NA)
    )
output %>% 
    complete(treatment, gender)
```

2.  Read the dataset available at "https://raw.githubusercontent.com/JaneWall/data_STAT412612/master/weather.csv" as weather. 

Use "pivot_longer()" to to put the days all in one column, then use "pivot_wider" to separate tmax and tmin into separate columns. Print the summary of the final resulting dataset.

```{r}
weather <- read_csv("./weather.csv")

weather %>% 
    pivot_longer(
        cols = starts_with("d"),
        names_to = "day",
        names_prefix = "d",
        values_drop_na = TRUE
    ) %>% 
        pivot_wider(
            names_from = "element",
            values_from = "value"
        ) %>% 
            summary()
```

3.  Tidy the billboard dataset (built-in).

      a.  First gather up all the week entries into a row for each week for each song (where there is an entry)
      b.  Then, convert the week variable to a number and figure out the date corresponding to each week on the chart
      c.  Sort the data by artist, track and week. Here are what your first entries should be (formatting can be different)

```{r}
billboard %>% 
    pivot_longer(
        cols = starts_with("wk"),
        names_to = "week",
        names_prefix = "wk",
        names_transform = list(week = as.integer),
        values_drop_na = TRUE
    ) %>% 
        mutate(date = date.entered + weeks(week)) %>% 
            arrange(artist, track, week)
```


4.  Load the built in "anscombe" data frame and use "pivot_longer()" to separate all the x and y columns and categorize them into 4 sets.

```{r}
anscombe %>% 
     pivot_longer( 
         cols = starts_with(c("x","y")),
         names_to = c(".value", "set"),
         names_pattern = "(.)(.)"
     )
```

5.  As explained in the video load and tidy the built in world_bank_pop data frame.

```{r}
world_bank_pop %>% 
    pivot_longer(
        cols = `2000`:`2017`,
        names_to = "year",
        values_to = "value",
        values_drop_na = TRUE
    ) %>% 
        separate_wider_regex(
            cols = indicator,
            patterns = c("^.*[:punct:]", 
                         area = ".*", 
                         "[:punct:]", 
                         variable = ".*$")
        ) %>% 
            pivot_wider(
                names_from = "variable",
                values_from = "value"
            )
```

