---
title: "Practice Tidying Data Lab"
author: "Anthony Tetreault"
date: '`r Sys.Date()`'
output: pdf_document
---

##### Libraries
Load the tidyverse library
```{r}
library(tidyverse)
```

##### Question 1. 
The following built-in datasets are not tidy. For each one, describe why it is not tidy, write out what the first five entries would look like once it is in a tidy format, and then tidy the dataset 

a.relig_income
b.billboard
c.us_rent_income

```{r}
# relig_income
# This dataset is not tidy because the columns are values of a variable, income, not each variable has its own column.
relig_income %>% 
    pivot_longer(
        cols = c(`<$10k`, `$10-20k`, `$20-30k`, 
                 `$30-40k`, `$40-50k`, `$50-75k`, 
                 `$75-100k`, `$100-150k`, `>150k`, 
                 `Don't know/refused`),
        names_to = "income",
        values_to = "count"
    )

# billboard
# This dataset is not tidy because the columns are values of a variable, weeks, not each variable has its own column.
billboard %>% 
    pivot_longer(
        cols = starts_with("wk"),
        names_to = "week",
        names_prefix = "wk",
        names_transform = list(week = as.integer),
        values_drop_na = TRUE
    ) %>% 
        mutate(date = date.entered + weeks(week)) %>% 
            arrange(artist, track, week)

# us_rent_income
# This dataset is not tidy because the column variable contains multiple variables in it (income, rent) 
# and each variable should have its own column.
us_rent_income %>% 
    pivot_wider(
        id_cols = c("GEOID", "NAME"),
        names_from = "variable",
        values_from = c("estimate", "moe")
    )

```

##### Question 2.
2.Use "pivot_longer" to tidy the built-in table4b dataset

```{r question-2-response}
table4b %>% 
    pivot_longer(
        cols = c(`1999`, `2000`),
        names_to = "year",
        values_to = "population"
    )

```

##### Question 3. 
3.Import and tidy the monkeymen dataset. The cell values represent identification accuracy of some objects (in percent of 20 trials).

```{r question-3-response}
monkeymen <- read_csv("./monkeymem.csv")

monkeymen %>% 
    pivot_longer( # Collect years into one column, so all columns have one variable
        cols = starts_with("Week"),
        names_to = "Week",
        names_prefix = "Week",
        values_to = "Percent"
    )

```

##### Question 4. 
4. As explained in the lecture video load and tidy the built in world_bank_pop data frame

```{r question-4-response}
world_bank_pop %>% 
    pivot_longer( # Collect years into one column
        cols = `2000`:`2017`,
        names_to = "year",
        values_to = "value",
        values_drop_na = TRUE
    ) %>% 
        separate_wider_regex( # Expand indicator column to capture area and variable
            cols = indicator,
            patterns = c("^.*[:punct:]", # SP.
                         area = ".*", # URB
                         "[:punct:]", # .
                         variable = ".*$") # TOTL/GROW
        ) %>% 
            pivot_wider( # Expand variable column to TOTL and GROW columns
                names_from = "variable",
                values_from = "value"
            )

```
