---
title: "Untitled"
author: "Anthony Tetreault"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ISLR)
```

1.  How would you write the following expression? Use writeLines() to see the raw result.
She said: "Let's all go!" \never mind! I said "I don't want to."

```{r include = FALSE}
writeLines("She said: \"Let's all go!\" \\never mind! I said \"I don't want to.\"")
```

2.  Use str_trim() and str_squish() to replace the string: 
"   Help me to      get rid    of extra spaces?      "
```{r}
x <- "   Help me to      get rid    of extra spaces?      "
# with the following strings: 

# "Help me to      get rid    of extra spaces?      "
str_trim(x, side = "left")
```


```{r}
# "   Help me to      get rid    of extra spaces?"
str_trim(x, side = "right")
```


```{r}
# "Help me to      get rid    of extra spaces?"
str_trim(x)
```


```{r}
# "Help me to get rid of extra spaces?"
str_squish(x)
```

3.  Using the words provided in stringr::words, create regular expressions that find all words that:
```{r}
# a. Start with "y"
str_view(words, "^y.*")
```

```{r}
# b. End with "x"
str_view(words, ".*x$")
```

```{r}
# c. End with "ed", but not with "eed"
str_view(words, "^.*[^e]ed$")
```

```{r}
# d. Start or end with "x"
str_view(words, "(^x)|(x$)")
```

```{r}
# e. Contains three or more vowels in a row
str_view(words, "[aeiou]{3,}")
```

```{r}
# f. Starts and ends with the same character
str_view(words, "^(.).*\\1$")
```

```{r}
# g. Contains at least 4 of the 5 vowels
vowel80pct <- c("^(?:(?=.*e)(?=.*i)(?=.*o)(?=.*u)|(?=.*a)(?=.*i)(?=.*o)(?=.*u)|(?=.*a)(?=.*e)(?=.*o)(?=.*u)|(?=.*a)(?=.*e)(?=.*i)(?=.*u)|(?=.*a)(?=.*e)(?=.*i)(?=.*o)).*$")
str_view(words, vowel80pct)

```

4.   Using the sentences provided in stringr::sentences, modify the regex so it no longer picks up the word flickered as a color (which was the case in the lecture video).
```{r}
colors <- c("^red", "orange", "yellow", "green", "blue", "purple$")
color_match <- str_c(colors, collapse = "|")
has_color <- str_subset(sentences, color_match)
matches <- str_extract(has_color, color_match)
more <- sentences[str_count(sentences, color_match) == 1]
# a.  Show all the sentences that contain at least one color with the word highlighted using str_view_all()
str_view(more, color_match)
# b.  Use str_extract_all() to produce an array with the colors that appear multiple times in a sentence.
str_extract_all(more, color_match, simplify = TRUE)

```

5.  Load the ISLR package. Then take names <- row.names(College).

```{r}
names <- row.names(College)
# a. Show the first twenty rows that result if you split Names at each space into 3 parts
names %>% 
    str_split(boundary("word"), 3, simplify = TRUE) %>% 
    head(20)
# b. Do they same by splitting the first 20 rows on every word boundary.
names %>% 
    str_split(boundary("word"), simplify = TRUE) %>% 
    head(20)
```
      
      