---
title: "Rmd Assignment"
author: "Anthony Tetreault"
date: "`r Sys.Date()`"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
```

# About Me

## Current Fave Artists w/ Albums

* **Aesop Rock**
  + _Impossible Kid_
  + *Garbology*
* __La Luz__
  + *News of the Universe*
  + _Floating Features_

## 5 States I Have Lived In
### __Ordered__ by _preference_

1. **Oregon**
1. South Carolina
1. Connecticut
1. New York
1. *Florida*

## Echo makes it look like magic!

```{r echo=FALSE}
x <- 6
y <- 7
x*y
```

## Tippecanoe and Tyler Too

1st year in   | Office        | Name
------------- | ------------- | ----------------------
1841          | President     | William Henry Harrison
1841          | President     | John Tyler
1976          | DCI           | George H.W. Bush
1981          | V. President  | George H.W. Bush
1989          | President     | George H.W. Bush
2019          | Representative| Rashida Tlaib

## A Plot

```{r}
ggplot(mpg, mapping=aes(displ, hwy, colour = class)) + geom_point() +  
  xlab("Engine Displacement") + ylab("Highway MPG")
```


#### Session Info
```{r}
sessionInfo()
```

